---
title: "Factors Associated with Laboratory
Sierra-Leone, West Africa:"
author: "Ojiambo Ivan"
date: "`r Sys.Date()`"
output:
    prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading Libraries----

```{r include=FALSE}
library(tidyverse) 
library(lme4) 
library(sjPlot) 
library(ggplot2) 
library(lmerTest) 
library(nlme) 
library(broom.mixed) 
library(performance) 
library(see) 
library(ineq) 
library(ggeffects) 
library(brms) 
library(sf) 
library(emmeans) 
library(readxl) 
library(dplyr) 
library(naniar) 
library(VIM) 
library(visdat)
library(stringr)
library(broom) 
library(pscl) 
library(here)
library(janitor)
library(ResourceSelection)
```

`Loading study Ebola data---

```{r echo=FALSE, message=FALSE}
ebola<-read_csv(here("data/ebola_sierra_leone.csv"))

View(ebola)

```

**CLEANING EBOLA_DATA----**

clean column 

```{r echo=FALSE, message=FALSE}

ebola<-clean_names(ebola) 
names(ebola) 
```


**Checking of missing data--**

```{r echo=FALSE, message=FALSE}
colSums(is.na(ebola)) 
colMeans(is.na(ebola))*100 
vis_miss(ebola)

```

**checking Duplicates---**

```{r echo=FALSE}
sum(duplicated(ebola$id)) 
```

**cleaning_age----**

```{r echo=FALSE}
summary(ebola$age)

ebola$age<-as.numeric(ebola$age) 
ebola<-ebola %>%  
filter(age>=0 & age<=100)

ebola$age[is.na(ebola$age)]<-median(ebola$age,na.rm = TRUE) 
summary(ebola$age) 
hist(ebola$age)

```

**cleaning_sex---** 

```{r echo=FALSE}
ebola$sex<-tolower(ebola$sex) 
ebola$sex<-recode(ebola$sex, 
"m"="male", 
"f"="female") 
ebola$sex<-factor(ebola$sex)
ebola$sex
```


**cleaning_status----** 

```{r echo=FALSE}
unique(ebola$status) 
ebola$status<-tolower(ebola$status) 
ebola$status<-factor(ebola$status, 
levels = c("suspected","confirmed")) 
```

**cleaning_districts----**

```{r echo=FALSE}
unique(ebola$district) 
ebola$district<-tolower(ebola$district) 
ebola$district<-trimws(ebola$district) 
ebola$district<-factor(ebola$district) 
```

**dates_cleaning---**

*date of onset*

```{r echo=FALSE}
library(lubridate) 
head(ebola$date_of_onset) 
ebola$date_of_onset<-ymd(ebola$date_of_onset)
```

*date of the sample*

```{r echo=FALSE}
head(ebola$date_of_sample) 
ebola$date_of_sample<-ymd(ebola$date_of_sample) 
```


**sample data before onset data---**

```{r echo=FALSE}
ebola %>%  
  filter(date_of_sample<date_of_onset) 
ebola<-ebola %>%  
  filter(date_of_sample>=date_of_onset) 
```
 
**delay variables---**

```{r echo=FALSE}
ebola$delay_days<-as.numeric( 
  ebola$date_of_sample - ebola$date_of_onset 
  )
ebola$delay_days<-as.data.frame(ebola$delay_days)
reactable::reactable(ebola$delay_days)

summary(ebola$delay_days) 
```


**Creating Age_group---**

```{r echo=FALSE}
min(ebola$age) 
max(ebola$age) 
ebola$age_group<-cut(ebola$age, 
                breaks = c(0,14,24,34,44,100), 
                labels = c("0-14","15-24","25-34","35-44","45+"), 
                right = TRUE) 
```

The minimum age was **`r min(ebola$age) `** and maximum age was **`r  max(ebola$age) `**

 
**structure check---**

```{r echo=FALSE}
str(ebola) 
ebola<-as.data.frame(summary(ebola))
reactable::reactable(ebola)
```
 
 
**ANALYSIS OF FACTORS ASSOCIATED WITH CONFIRMED EBOLA---- **

**Descriptive analysis**

```{r echo=FALSE}

cleaned_ebola <- read_csv(here("data/cleaned_ebola_data.csv"))
 
str(cleaned_ebola) 
summary<-as.data.frame(summary(cleaned_ebola))
reactable::reactable(summary) 
```

**Total number of cases after cleaning--**

```{r echo=FALSE}
number_of_cases<-nrow(cleaned_ebola) 
number_of_cases
```
 
**frquency destribution of status---**

```{r echo=FALSE}
status_tab<-table(cleaned_ebola$status)
status_tab
```

**percentage of status--**

```{r echo=FALSE}
percentage_status<-prop.table(status_tab)*100

percentage_status<-as.data.frame(percentage_status)
reactable::reactable(percentage_status)

```

 
**frequency_destribution of sex---**

```{r echo=FALSE}
sex_tab<-table(cleaned_ebola$sex) 
percentage_sex<-prop.table(sex_tab)*100 
percentage_sex<-as.data.frame(percentage_sex) 
reactable::reactable(percentage_sex)
```
 
**Cross_tabulation of Sex Vs Status---**

```{r echo=FALSE}
cross_tab_sex<-table(cleaned_ebola$sex,cleaned_ebola$status) 
cross_tab_sex 
percentage_cross_tabsex<-prop.table((cross_tab_sex),1)*100 
percentage_cross_tabsex<-as.data.frame(percentage_cross_tabsex)
reactable::reactable(percentage_cross_tabsex)
```
 
**Descriptive statistics for Age--**


```{r echo=FALSE}
cleaned_ebola |> 
  summarise(
    Mean=mean(age,na.rm=T) ,
    SD=sd(age) ,
    Meadian=median(age,na.rm=T), 
   IQR=IQR(age,na.rm=T) 

  )

```

**Age_group--**

```{r echo=FALSE}
age_group_tab<-table(cleaned_ebola$age_group) 
percentage_age_group<-prop.table(age_group_tab)*100 
 
age_group_tab 
percentage_age_group 
``` 
 
**Age Group Vs status---**

```{r echo=FALSE}
crosstab_age<-table(cleaned_ebola$age_group, cleaned_ebola$status) 
percentage_cross_age<-prop.table((crosstab_age),1)*100 
crosstab_age 
percentage_cross_age 
```
 
**District destribution--**

```{r echo=FALSE}
district_tab<-table(cleaned_ebola$district) 
district_tab 
 
sorting<-sort(district_tab,decreasing = TRUE) 
sorting 
percentage<-prop.table((district_tab))*100 
```

**District Vs Status---**

```{r echo=FALSE}
district_crosstab<-table(cleaned_ebola$district,cleaned_ebola$status)  
  
reactable::reactable(district_crosstab)
```

**Delay Days Summary---**

```{r echo=FALSE}
summary(cleaned_ebola$delay_days) 
mean(cleaned_ebola$delay_days) 
median(cleaned_ebola$delay_days) 
```

**Epidermiological Week**

```{r echo=FALSE}
cleaned_ebola$epi_week<- isoweek(cleaned_ebola$date_of_onset) 


weekly_cases<-cleaned_ebola %>%  
group_by(epi_week) %>%  
summarise(cases=n()) 
reactable::reactable(weekly_cases)
```
 
**Epidemic curve--**

```{r echo=FALSE, message=FALSE}
p<-ggplot(weekly_cases, 
       aes(x=epi_week,y=cases))+ 
  geom_col()+ 
  labs(title = "Epidemic Curve of Cases", 
       x="Epidemiological Week", 
       y="Number of cases") 
p 

```

**Epidemic curve by status--**

```{r echo=FALSE, message=FALSE}
weekly_status<-cleaned_ebola %>%  
  group_by(epi_week,status) %>%  
  summarise(cases = n()) 
weekly_status_plot<-ggplot(weekly_status, 
       aes(x=epi_week, 
           y=cases, 
           fill = status))+ 
  geom_col(position = "stack")+ 
  labs(title = "Weekly Ebola Cases by Status", 
       x="Epidemiological Week", 
       y="Number of cases") 
weekly_status_plot 
``` 
 
**Bar-chart of sex Distribution---**

```{r echo=FALSE}
sex_barchart<-ggplot(cleaned_ebola, 
       aes(x=sex,))+ 
  geom_bar()+ 
  labs(title = "Distribution of cases by Sex", 
       x="Sex", 
       y="Frequecy") 
sex_barchart 
```


**Bar_chart of Age distribution**

```{r echo=FALSE}
Age_group_barchart<-ggplot(cleaned_ebola, 
                     aes(x=age_group))+ 
  geom_bar()+ 
  labs(title = "Distribution of Cases by Age Group", 
       x="Age Group", 
       y="Frequency") 
Age_group_barchart 
```
 
**Boxplot of Age by status**

```{r echo=FALSE}
boxplot_age_Vs_status<-ggplot(cleaned_ebola, 
       aes(x=status, 
           y=age))+ 
  geom_boxplot()+ 
  labs(title = "Age Distribution by Ebola Status", 
       x="Status", 
       y="Age") 
boxplot_age_Vs_status 
 
```

**Frequency Tables with percentages Together--**

```{r echo=FALSE}
sex_table<-cleaned_ebola %>%  
  count(sex) %>%  
  mutate(percent=round(n/sum(n)*100, 1)) 
reactable::reactable(sex_table) 
```

**Status Proportion by District---**

```{r echo=FALSE, message=FALSE}
district_status<-cleaned_ebola %>%  
  group_by(district,status) %>%  
  summarise(n=n()) %>%  
  mutate(percent=round(n/sum(n)*100, 1)) 
reactable::reactable(district_status)
```

 
**Chi-Square Analysis---- **

**chi-square test: Sex Vs Status**

```{r echo=FALSE, message=FALSE}
sex_status<-table(cleaned_ebola$sex,cleaned_ebola$status) 
reactable::reactable(sex_status)

chisq.test(sex_status) 
 
reactable::reactable(prop.table(sex_status,1)*100)
```

**chi-square test:Age_group Vs status**

```{r echo=FALSE, message=FALSE, warning=FALSE}
age_status<-table(cleaned_ebola$age_group,cleaned_ebola$status) 
reactable::reactable(age_status)

chisq.test(age_status) 
fisher.test(age_status) 


reactable::reactable(prop.table(age_status,1)*100)
```


**chi-square test:Districts Vs Status**

```{r echo=FALSE, message=FALSE, warning=FALSE}
cleaned_ebola$district_group<-ifelse( 
cleaned_ebola$district %in% c("western urban","bo","kono","port loko","kambia"), 
"other District", 
as.character(cleaned_ebola$district) 
) 
cleaned_ebola$district_group<-factor(cleaned_ebola$district_group) 
district_status<-table(cleaned_ebola$district_group,cleaned_ebola$status) 
reactable::reactable(district_status)

chisq.test(district_status)
reactable::reactable(chisq.test(district_status)$expected)


fisher.test(district_status)
```

**Chi-Square Test workflow**

**Sex**

```{r echo=FALSE, message=FALSE, warning=FALSE}
chisq.test(table(cleaned_ebola$sex,cleaned_ebola$status))
```

**Age_group**

```{r echo=FALSE, message=FALSE, warning=FALSE}
chisq.test(table(cleaned_ebola$age_group,cleaned_ebola$status)) 
fisher.test(table(cleaned_ebola$age_group,cleaned_ebola$status))
```

**District**

```{r echo=FALSE, message=FALSE, warning=FALSE}
chisq.test(table(cleaned_ebola$district,cleaned_ebola$status)) 
fisher.test(table(cleaned_ebola$district,cleaned_ebola$status)) 
```

**summary of sex table**

```{r echo=FALSE, message=FALSE}
sex_summary<-cleaned_ebola %>%  
group_by(sex,status) %>%  
summarise(n=n()) %>%  
mutate(percent=round(n/sum(n)*100, 1)) 
reactable::reactable(sex_summary)
 
```

**Visualization of association----**

**stacked bar-chart**

*sex Vs status* 

```{r echo=FALSE, message=FALSE}
visual_sexVs_status<-ggplot(cleaned_ebola, 
       aes(x=sex, 
           fill = status))+ 
  geom_bar(position = "fill")+ 
  labs(title = "Sex and Ebola Status", 
       x="Sex", 
       y="Proportion") 
visual_sexVs_status 
```
 
**Age group Vs status** 

```{r echo=FALSE, message=FALSE}
agegroupVs_status<-ggplot(cleaned_ebola, 
       aes(x=age_group, 
           fill = status))+ 
  geom_bar(position = "fill")+ 
  labs(title = "Age Group and Ebola Status", 
       x="Age Group", 
       y="Proportion") 
agegroupVs_status 
 
```

**Binary Logistic Regression---- **

```{r echo=FALSE, message=FALSE}
str(cleaned_ebola$status) 
cleaned_ebola$status<-factor(cleaned_ebola$status, 
levels = c("suspected","confirmed") 
) 
levels(cleaned_ebola$status) 
str(cleaned_ebola$status)
```

**Reference category checking**

```{r echo=FALSE, message=FALSE}
cleaned_ebola$sex<-factor(cleaned_ebola$sex) 
levels(cleaned_ebola$sex) 
cleaned_ebola$age_group<-factor(cleaned_ebola$age_group) 
levels(cleaned_ebola$age_group) 
cleaned_ebola$district<-factor(cleaned_ebola$district_group) 
levels(cleaned_ebola$district_group)
```

**Setting of reference category**

```{r echo=FALSE, message=FALSE}
cleaned_ebola$sex<-relevel(cleaned_ebola$sex,ref = "female") 
cleaned_ebola$age_group<-relevel(cleaned_ebola$age_group,ref = "0-14") 
cleaned_ebola$district<-relevel(cleaned_ebola$district,ref = "kailahun")

```

**Crude Logistic Regression---**

**sex only**

```{r echo=FALSE, message=FALSE}
model_sex<-glm( 
status~sex, 
data = cleaned_ebola, 
family = binomial 
) 
summary(model_sex)

exp(coef(model_sex))

reactable::reactable(exp(confint(model_sex))) 
```

**logistic regression for Age Group**

```{r echo=FALSE, message=FALSE}
model_age<-glm( 
status~age_group, 
data=cleaned_ebola, 
family = binomial 
) 
summary(model_age)

exp(coef(model_age)) 
 
reactable::reactable(exp(confint(model_age)))
```

**logistic for District**

```{r echo=FALSE, message=FALSE}
model_district<-glm( 
status~district, 
data=cleaned_ebola, 
family = binomial 
) 
summary(model_district) 

exp(coef(model_district)) 

reactable::reactable(exp(confint(model_district)))
```

**FUll Multivariable Logistic Regression----**

```{r echo=FALSE, message=FALSE}
full_model<-glm( 
status~age_group + sex + district, 
data = cleaned_ebola, 
family = binomial 
) 
summary(full_model)
```

**Adjusted Odds ratio(AOR)--**

```{r echo=FALSE, message=FALSE}
exp(coef(full_model)) 

reactable::reactable(exp(confint(full_model))) #95% confidence Interval 
 

results<-tidy(full_model, 
exponentiate = TRUE, 
conf.int = TRUE) 
reactable::reactable(results) 
```


**Model Fit Assessment----**

```{r echo=FALSE, message=FALSE}

pR2(full_model)
```

**Hosmer_lemeshow Goodness of fit---- **

```{r echo=FALSE, message=FALSE}
hoslem.test( 
as.numeric(cleaned_ebola$status) - 1, 
fitted(full_model) 
) 
```

**Predict Probability---- **

```{r echo=FALSE, message=TRUE}
cleaned_ebola$predicted_prob<-predict(full_model, 
type = "response") 
reactable::reactable(as.data.frame(cleaned_ebola$predicted_prob)) 
head(cleaned_ebola$predicted_prob)

```







